---
title: "monthly_bus_delays"
output: html_document
date: "2023-11-30"
---

# Set up python enviroment in R & import pkgs
```{r reticulate setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(ggiraph)
library(councildown)
knitr::opts_chunk$set(echo = TRUE)
require("knitr")
opts_knit$set(root.dir = "~/Desktop/Data_Projects/schoolBusDelays2022")
library(reticulate)
#use_condaenv("r-reticulate")
scipy <- import("scipy")
dt <- import("datetime")
pd <- import("pandas")
np <- import("numpy")
```

# load pkgs
```{python import python pkgs}
import pandas as pd
import numpy as np
from scipy.stats import pearsonr
from scipy.stats import norm
import re
import requests
import calendar
import copy
```
# Data Cleaning
 
## read in calendar dates files
```{python read in data, echo=FALSE}
data = pd.read_csv('../data/input/modified/DOE School Calendar Dates_edited.csv').drop(columns = ['Unnamed: 3'])
data = data.replace('–', '-') 
```

## clean dates, function to get range dates sorted out
```{python clean dates, echo=FALSE}
# Data start date should be Sept. 7, 2017

# Have to preprocess date ranges
# Split into start and end dates. If there is a hyphen, 
# take the two numbers on either side of it and use one as start date and one as an end date

# 27-30 Dec 2022

# For this to work, replace all '‚Äì' (big hyphen code) in csv with hyphens '-' 

start_dates, end_dates = [], []
for i in data['Dates']:
    if ('-' not in i):
        start_dates.append(pd.to_datetime(i))
        end_dates.append(pd.to_datetime(i))
    else:
        x = i.split(' ')
        start_dates.append(pd.to_datetime(' '.join([k.split('-')[0] if ('-' in k) else k for k in x])))
        end_dates.append(pd.to_datetime(' '.join([k.split('-')[1] if ('-' in k) else k for k in x])))
        
data['Start Date'] = start_dates
data['End Date'] = end_dates 
```

## define and run school calendar yr function
```{python school yr func}
def get_year(school_year, month):
    if int(month) < 9:
        return int(school_year[:4]) + 1
    else:
        return int(school_year[:4]) 
```

```{python apply func}
data['End Date'] = pd.to_datetime(data['End Date'].dt.month.astype(str) + "/" + data['End Date'].dt.day.astype(str) + "/" + data.apply( (lambda x: get_year(x['School Year'], x['End Date'].month)), axis = 1).astype(str))

data['Start Date'] = pd.to_datetime(data['Start Date'].dt.month.astype(str) + "/" + data['Start Date'].dt.day.astype(str) + "/" + data.apply( (lambda x: get_year(x['School Year'], x['Start Date'].month)), axis = 1).astype(str))
```

## account for covid
```{python covid func}
# Adding covid shutdown 3/16/2020 - 9/13/2021

covid = {'School Year': '2021-2022', 
         'Dates': '3/16/2020 - 9/13/2021', 
         'Reason': 'COVID', 
         'Start Date': pd.to_datetime('3/16/2020'), 
         'End Date': pd.to_datetime('9/13/2021')
        }

covid_df = pd.DataFrame([covid])
data = pd.concat([data, covid_df], ignore_index=True)
```

## reference safe keeps in R
```{r}
r <- py$data
```

### Non-holiday closure dates, students not in attendance
```{python non-holidays closed}
# Sort reasons
data['Closure'] = data['Reason'].apply( lambda x: 1 if  ( ("closed" in x) | 
                                       ("no students" in x) | 
                                       ("9-12 schools only" in x) | 
                                       ("remote" in x) | 
                                       ("hours early" in x) | 
                                       ("do not attend" in x) | 
                                       ("CLOSED" in x) | 
                                       ("Clerical Day" in x) | 
                                       ("staff development" in x) | 
                                       ("Students not in attendance" in x) | 
                                       ("Students will not" in x) | 
                                       ("will not be in attendance" in x) | 
                                       ("COVID" in x) | 
                                       ("Summer Recess" in x) ) else 0 )
```

########

# Analysis 

## read in school enrollment data
```{python enrollment nums}
# Enrollment numbers
# since enrollment #s go back to 2017, we can only use school bus delay #s from 2017 to normalize the data
# https://infohub.nyced.org/reports/school-quality/information-and-data-overview
enrollment_count = {
    2017: 1135334,
    2018: 1126501,
    2019: 1131868,
    2020: 1094138,
    2021: 1058888,
    2022: 1047895,
    2023: 1047895*0.989 #This is an based on the 1.1% drop in school population from last year. Per info provided by DOE
}
```

# save to data folder
```{r}
enroll<- plyr::ldply(py$enrollment_count, data.frame)
names(enroll)<- c("year", "enrollment")

write.csv(enroll, "../data/input/raw/enrollment_nums.csv", row.names=F)
```

## read in school bus delay data
```{python school bus delays data}
#data_orig = pd.read_json("https://data.cityofnewyork.us/resource/ez4e-fazm.json?$limit=99999999&$where=school_year%20not%20in(%272015-2016%27,%20%272016-2017%27)")

#data_orig['occurred_on'] = pd.to_datetime(data_orig['occurred_on'])

# using this to filter out Feb dates
#data_orig = data_orig[data_orig['occurred_on'] < '2024-02-01']
```

# code above was not working, so using altnerative
```{python}
url = "https://data.cityofnewyork.us/resource/ez4e-fazm.json?$limit=99999999&$where=school_year%20not%20in(%272015-2016%27,%20%272016-2017%27)"
response = requests.get(url, verify=True)
data_orig = pd.read_json(response.text)

data_orig['occurred_on'] = pd.to_datetime(data_orig['occurred_on'])

# using this to filter to dates of interest
data_orig = data_orig[data_orig['occurred_on'] < '2024-07-01']
```

## checking dates
```{r}
View(py$data_orig)
```

## reference safe keeps in R
```{r}
r1 <- py$data_orig
```

## Function that cleans how_long_delayed column
```{python func for clean delay times}
def clean_delay_times(x):
    a = x
    x = str(x).lower().rstrip().replace('1 1/2', '1 2').replace('11/2', '1 2').\
    replace('1/2', '0 1').replace('half', '0 1').replace('to', '-')
    units = re.findall(r'[a-z ]+', x)
    time = re.split(r'[- /,]', ''.join(re.findall(r'[0-9 .,/-]+', x)) )
    time = [i for i in time if not((re.search('[.]+',i) !=None)|(re.search('[ ]+',i) !=None)|(i==''))]
    if time:
        time = np.nanmean([float(i) for i in time])
    else:
        time = 0
    
    if any(i in ''.join(units) for i in ['h', 'r']) and ( time <= 5 ):
        return time*60
    elif any(i in ''.join(units) for i in ['m', 'n']) | (x == ''):
        return time
    else:
        return 0
```

```{python clean how long delay col}

data_orig['delay_time'] = data_orig.how_long_delayed.astype(str).apply(clean_delay_times)
```

```{python}

days_closed = data[data['Closure'] == 1]

###########Faster Version####################
data_orig['closed'] = 1

for row, val in days_closed.iterrows():
    a = val['Start Date']
    b = val['End Date'] + pd.DateOffset(1)
    data_orig.loc[:,('closed')] = data_orig.loc[:,('closed')]*(~data_orig.loc[:,('occurred_on')].apply(
        lambda x: (x > a) & (x < b)))

data_orig = data_orig[data_orig['closed'] == 1]
#############################################
# Eliminate weekends

data_orig = data_orig[data_orig['occurred_on'].dt.dayofweek <= 4]
```

# downloading
```{python}
data_orig.to_csv("../data/output/filtered_weekends_vacation_covid_delays.csv")
```

# Generating Visuals

```{r}
x <- read_csv("../data/output/filtered_weekends_vacation_covid_delays.csv")
# read in enrollment nums
enroll <- read_csv("../data/input/raw/enrollment_nums.csv") %>% 
  mutate(year = as.numeric(year))
```

# 01 Total Delays per Month ----
```{r}
# data prep 
delays_monyr <- x %>% 
  mutate(day=day(occurred_on), 
         month=month(occurred_on), 
         year=year(occurred_on)) %>% 
  group_by(year, month) %>% 
  summarize(count=n(), 
            average_month=mean(delay_time)) %>% 
  mutate(month_char=month.abb[month], 
         my=factor(month_char, levels= unique(month_char)),
         monyr=ym(paste(year,month, sep=" ")),
         School_Year = case_when(monyr >='2017-09-01' & 
                                   monyr<='2018-06-01' ~ '2017-2018',
                                 monyr >='2018-09-01' & 
                                   monyr<='2019-06-01' ~ '2018-2019',
                                 monyr >='2019-09-01' & 
                                   monyr<='2020-06-01' ~ '2019-2020',
                                 monyr >='2020-09-01' & 
                                   monyr<='2021-06-01' ~ '2020-2021',
                                 monyr >='2021-09-01' & 
                                   monyr<='2022-06-01' ~ '2021-2022',
                                 monyr >='2022-09-01' & 
                                   monyr<='2023-06-01' ~ '2022-2023',
                                 monyr >='2023-09-01' & 
                                   monyr<='2024-06-01' ~ '2023-2024'
                                ))
# filtering out N/A values
delays_monyr = delays_monyr %>%
  filter(!is.na(School_Year))

# plot

plot <-   
  ggplot(data = delays_monyr, 
         aes(x = my,  y = count)) +
  geom_line(aes(color=School_Year, group=School_Year), alpha=1) +
  geom_point_interactive(size=1.5, alpha=0.85,
                         aes(color=School_Year, group=School_Year),
                         tooltip = paste(delays_monyr$month_char,
                                         delays_monyr$year, " :", 
                                scales::comma(delays_monyr$count),
                                         " delays")) + 
  scale_color_nycc(palette = "main", reverse = T) +
  # geom_smooth(aes(group=1), se=F, size=0.75,
  #                         alpha =.5, color="#23417D") +
  scale_y_continuous(breaks = seq(0,
                                  max(delays_monyr$count),
                                  1000),
                     labels = scales::comma(seq(0,
                                  max(delays_monyr$count),
                                  1000))) +
  labs(title="Number of Delays Over Time", 
       x="School Year Calendar Months",  
       y="Number of Delays", color="SY") +
  theme_nycc() 

tooltip_css <- "background-color:#CACACA;"

plot_interactive <- girafe(ggobj = plot,   
                           width_svg = 9,
                           height_svg = 5, 
                           options = list(
                             opts_tooltip(css = tooltip_css)
                           )
)

plot_interactive

htmlwidgets::saveWidget(plot_interactive, file = "../visuals/num_monthly_delays.html")
```

# 02 Total Delays per Month (as percent of total trips) ----
```{r}
# Adding column for total monthly trips 
trips_per_month = read_csv('../data/output/total_bus_trips_per_month.csv')
trips_per_month$Date <- as.Date(trips_per_month$Date)
trips_per_month <- trips_per_month %>% rename(total_trips = 'Total Count of Trips')

# Extract 'year' and 'month' from date columns in both data frames
delays_monyr <- delays_monyr %>%
  mutate(year = year(monyr), month = month(monyr))

trips_per_month <- trips_per_month %>%
  mutate(year = year(Date), month = month(Date))

# Merge based on 'year' and 'month'
per_delays_monyr <- merge(
  delays_monyr, 
  trips_per_month %>% select(year, month, total_trips), 
  by = c("year", "month")
) %>% rename(total_delays = 'count')

# Find percent 
per_delays_monyr['delay_per_of_total_trips'] = 100*per_delays_monyr['total_delays'] / per_delays_monyr['total_trips']
per_delays_monyr['delay_per_of_total_trips'] <- round(per_delays_monyr['delay_per_of_total_trips'], 2)

# Dropping N/A
per_delays_monyr <- per_delays_monyr %>%
  filter(!is.na(delay_per_of_total_trips))

# Plot

plot <-   
  ggplot(data = per_delays_monyr, 
         aes(x = my,  y = delay_per_of_total_trips)) +
  geom_line(aes(color=School_Year, group=School_Year), alpha=1) +
  geom_point_interactive(size=1.5, alpha=0.85,
                         aes(color=School_Year, group=School_Year),
                         tooltip = paste(per_delays_monyr$month_char,
                             per_delays_monyr$year, ":", 
                             round(per_delays_monyr$delay_per_of_total_trips, 2), 
                             "% of all trips")) + 
  scale_color_nycc(palette = "main", reverse = T) +
  scale_y_continuous(breaks = seq(0,
                                  max(per_delays_monyr$delay_per_of_total_trips),
                                  1000),
                     labels = scales::comma(seq(0,
                                  max(per_delays_monyr$delay_per_of_total_trips),
                                  1000))) +
  labs(title="Delayed Trips As Percent of All Trips", 
       x="School Year Calendar Months",  
       y="Percent of All Trips", color="SY") +
  theme_nycc() 

tooltip_css <- "background-color:#CACACA;"

plot_interactive <- girafe(ggobj = plot,   
                           width_svg = 9,
                           height_svg = 5, 
                           options = list(
                             opts_tooltip(css = tooltip_css)
                           )
)

plot_interactive

htmlwidgets::saveWidget(plot_interactive, file = "../visuals/percent_monthly_delays.html")

selected_columns <- per_delays_monyr[, c("School_Year", "year", "month_char", "total_delays", "total_trips", "delay_per_of_total_trips")]
write.csv(selected_columns, "../data/output/percent_monthly_delays.csv", row.names = FALSE)
```
```{r}
View(per_delays_monyr)
```

# 03 Avg Delay Times per Month ----
```{r}
# same dataset is used

# plot

plot <-   
 delays_monyr %>% 
  ggplot( aes(x = my,  y = average_month)) +
  geom_line(aes(color=School_Year, group=School_Year), alpha=1) +
  geom_point_interactive(size=1.5, alpha=0.85,
                         aes(color=School_Year, group=School_Year),
                         tooltip = paste(delays_monyr$month_char,
                                         delays_monyr$year, " :",
                        scales::number(delays_monyr$average_month,
                                       accuracy = 1),
                                         " minutes") )+ 
  scale_color_nycc(palette = "main", reverse = T) +
  # geom_smooth(aes(group=1), se=F, size=0.75,
  #                         alpha =.5, color="#23417D") +
  scale_y_continuous(breaks = seq(0, max(delays_monyr$average_month),
                                  2),
                     labels = seq(0,max(delays_monyr$average_month),
                                                2)) +
  labs(title="Average Daily Delay Times", 
       x="School Year Calendar Months",  
       y="Minutes", color="SY") +
  theme(axis.text.x = element_text(angle = 0, hjust = 1, size = 9),
        legend.position = "top",
        axis.text.y = element_text(size = 9))

tooltip_css <- "background-color:#CACACA;"

plot_interactive <- girafe(ggobj = plot,   
                           width_svg = 9,
                           height_svg = 5, 
                           options = list(
                             opts_tooltip(css = tooltip_css)
                           )
)

plot_interactive

htmltools::save_html(plot_interactive,"../visuals/avg_monthly_delay_times.html")
```

# 04 Longest delays by reason -----
```{r}

# data prep
reasons <- x %>% 
  filter(reason!="")  %>% 
  mutate(day=day(occurred_on), 
         month=month(occurred_on), 
         year=year(occurred_on)) %>% 
  filter(year>=2021) %>% 
  group_by(reason) %>% 
  summarize(count=n(), 
            average_time=mean(delay_time)) %>% 
  top_n(5, wt = average_time)

# plot
plot <- 
  reasons %>% 
  ggplot(aes(x = reorder(reason,average_time), 
             y=average_time, fill = reason)) +
  geom_col_interactive(width = 0.6,
                      tooltip = 
                         paste("Reason:", reasons$reason, 
                               "<br>Average:", 
                               round(reasons$average_time,2))) +
  scale_fill_nycc(palette = "main") +
  coord_flip() +
  geom_text(show.legend = F, size = 3,
            label= paste0(round(reasons$average_time, 0), " min."), 
            nudge_x = 0, nudge_y = 4) +
  scale_y_continuous(expand = expansion(mult = c(0, .1))) +
  ylab("Average Minutes Delayed") + xlab("") +
  labs(title="",
       subtitle = "Longest Delays", 
       x="",  y="Average Minutes Delayed") +
  theme(axis.text.x = element_text(angle = 0, hjust = 1, size = 9),
        legend.position = "none",
        axis.text.y = element_text(size = 9))

tooltip_css <- "background-color:#CACACA;"

plot_interactive <- girafe(ggobj = plot,   
                           width_svg = 6,
                           height_svg = 5, 
                           options = list(
                             opts_tooltip(css = tooltip_css)
                           )
)

#htmltools::save_html(plot_interactive, "../visuals/longest_delays.html")
```

# 05 Most delays by reason -----
```{r}

# data prep
reasons <- x %>% 
  filter(reason!="")  %>% 
  mutate(day=day(occurred_on), 
         month=month(occurred_on), 
         year=year(occurred_on)) %>% 
  filter(year>=2021) %>% 
  group_by(reason) %>% 
  summarize(count=n()) %>% 
  mutate(percent=round(count/sum(count)*100,2)) %>% 
  top_n(5, wt = percent)


# plot
plot <- 
  reasons %>% 
  ggplot(aes(x = reorder(reason, percent), 
             y=count, fill = reason)) +
  geom_col_interactive(width = 0.6,
                       tooltip = 
            paste0("Reason: ", reasons$reason, 
                               "<br>Count: ", 
                      scales::comma(round(reasons$count,2)), 
                   "<br>Percent: ", 
                     scales::percent(reasons$percent, scale = 1))
                ) +
  scale_fill_nycc(palette = "main") +
  coord_flip() +
  geom_text(show.legend = F, size = 3,
            label= scales::comma(round(reasons$count,2)), 
            nudge_x = 0, hjust = -0.15) + #nudge_y not working for some reason
  scale_y_continuous(breaks = seq(0,max(reasons$count), 10000),
                     labels = scales::comma(
                       round(seq(0,max(reasons$count),10000),2)),
                     expand = expansion(mult = c(0, .1))) +
  labs(title="",
       subtitle = "Most Delays", 
       x="",  y="Number of Delays", color="SY") +
  theme(axis.text.x = element_text(angle = 0, hjust = 1, size = 9),
        legend.position = "none",
        axis.text.y = element_text(size = 9))

tooltip_css <- "background-color:#CACACA;"

plot_interactive <- girafe(ggobj = plot,   
                           width_svg = 6,
                           height_svg = 5, 
                           options = list(
                             opts_tooltip(css = tooltip_css)
                           )
)

#htmltools::save_html(plot_interactive, "../visuals/most_delays.html")
```

# 06 Reasons for delay over time ----
```{r}

# data prep
top_reasons <- x %>% 
  mutate(day=day(occurred_on), 
         month=month(occurred_on), 
         year=year(occurred_on)) %>% 
  filter(year>=2021) %>% 
  filter(reason!="") %>% 
  group_by(reason) %>% 
  summarize(count=n()) %>% 
  mutate(percent=round(count/sum(count)*100,3)) %>% 
  arrange(desc(count)) %>% top_n(5, wt=percent) %>% 
  select(reason) %>% unlist()

t <- x %>% 
  mutate(day=day(occurred_on), 
         month=month(occurred_on), 
         year=year(occurred_on)) %>% 
  filter(reason %in% top_reasons) %>% 
  group_by(year, month, reason) %>% 
  summarize(count=n(), 
            average_month=mean(delay_time)) %>% 
  
  mutate(reason = factor(reason, levels = top_reasons),
    month_char=month.abb[month], 
         my=factor(month_char, levels= unique(delays_monyr$month_char)),
    month=factor(month, levels= unique(delays_monyr$month)),
         monyr=ym(paste(year,month, sep=" ")),
         School_Year = case_when(monyr >='2017-09-01' & 
                                   monyr<='2018-06-01' ~ '2017-2018',
                                 monyr >='2018-09-01' & 
                                   monyr<='2019-06-01' ~ '2018-2019',
                                 monyr >='2019-09-01' & 
                                   monyr<='2020-06-01' ~ '2019-2020',
                                 monyr >='2020-09-01' & 
                                   monyr<='2021-06-01' ~ '2020-2021',
                                 monyr >='2021-09-01' & 
                                   monyr<='2022-06-01' ~ '2021-2022',
                                 monyr >='2022-09-01' & 
                                   monyr<='2023-06-01' ~ '2022-2023')) %>% 
  
  left_join(enroll, by="year") %>% #normalize to 2022 pop levels
  mutate(norm_count=round(count*1039828/enrollment,2)) 


plot <- t %>% 
  ggplot(aes(x = my,  y = norm_count)) +
  geom_line(aes(color=School_Year, group=School_Year), 
            alpha=1) +
  facet_wrap(~reason, ncol=5) +
  geom_point_interactive(size=1.5, alpha=0.85,
                         aes(color=School_Year, 
                             group=School_Year),
                         tooltip = paste(t$month_char,
                                         t$year, " :",
                                         scales::comma(t$norm_count,
                                                        accuracy = 1),
                                         " delays") )+ 
  scale_color_nycc(palette = "main", reverse = T) +
  scale_y_continuous(breaks = seq(0, max(t$norm_count),
                                  1000),
                     labels = scales::comma(seq(0,max(t$norm_count),
                                  1000))) +
  labs(title="", 
       x="School Year Calendar Months",  
       y="Number of Delays",
       color = "SY") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
         legend.position = "top",
         axis.text.y = element_text(size = 9))

tooltip_css <- "background-color:#CACACA;"

plot_interactive <- girafe(ggobj = plot,   
                           width_svg = 10,
                           height_svg = 5, 
                           options = list(
                             opts_tooltip(css = tooltip_css)
                           )
)

#htmltools::save_html(plot_interactive, "../visuals/reasons_num_delays.html")
```

# 07  SWD delays by delay type ----
```{r}

#read in new data that was previously missing
library(readr)
myData <- read_csv("../data/output/school_bus_delays_2022_updated.csv.zip")

# data prep
swd <- myData %>% 
  mutate(day=day(Occurred_On), 
         month=month(Occurred_On), 
         year=year(Occurred_On)) %>% 
  mutate(type=ifelse(grepl("General", Run_Type)==TRUE, 
                     "General Ed",
                     ifelse(grepl("Special", Run_Type)==TRUE,
                            "Special Ed","Other"))) %>% 
  group_by(year, month, type) %>% 
  summarize(count=n(), 
            average_month=mean(delay_time)) %>% 
  mutate(month_char=month.abb[month],
         my=factor(month_char, 
                   levels= unique(month_char))) %>%
  filter(type=="General Ed"| type=="Special Ed") %>% 
  mutate(type = factor(type, levels = c("Special Ed", "General Ed")),
         monyr=ym(paste(year,month, sep=" ")),
         School_Year = case_when(monyr >='2017-09-01' & 
                          monyr<='2018-06-01' ~ '2017-2018',
                        monyr >='2018-09-01' & 
                          monyr<='2019-06-01' ~ '2018-2019',
                        monyr >='2019-09-01' & 
                          monyr<='2020-06-01' ~ '2019-2020',
                        monyr >='2020-09-01' & 
                          monyr<='2021-06-01' ~ '2020-2021',
                        monyr >='2021-09-01' & 
                          monyr<='2022-06-01' ~ '2021-2022',
                        monyr >='2022-09-01' & 
                          monyr<='2023-06-01' ~ '2022-2023'))
# plot
plot <- swd %>% 
  ggplot(aes(x = my,  y = average_month)) +
  geom_line(aes(color=School_Year, group=School_Year), 
            alpha=1) +
  facet_wrap(~type, ncol=2) +
  geom_point_interactive(size=1.5, alpha=0.85,
                         aes(color=School_Year, 
                             group=School_Year),
             tooltip = paste(swd$month_char,
                             swd$year, " :",
                scales::number(swd$average_month,
                               accuracy = 1),
                 " minutes") )+ 
  scale_color_nycc(palette = "main", reverse = T) +
  scale_y_continuous(breaks = seq(0, max(swd$average_month),
                                  4),
                     labels = seq(0,max(swd$average_month),
                                  4)) +
  labs(title="", 
       x="School Year Calendar Months",  
       y="Average Daily Minutes",
       color="SY") +
  theme(axis.text.x = element_text(angle = 0, hjust = 1, size = 9),
        legend.position = "top",
        axis.text.y = element_text(size = 9))
  
  tooltip_css <- "background-color:#CACACA;"
  
  plot_interactive <- girafe(ggobj = plot,   
                             width_svg = 10,
                             height_svg = 5, 
                             options = list(
                               opts_tooltip(css = tooltip_css)
                             )
  )

#htmltools::save_html(plot_interactive, "../visuals/swd_delaytimes.html")
```




